---
title: "Predicting_Students_Dropout_and_Academic_Success"
author: "James Kibiro, Josh Balderas, Neelkumar Patel"
format: html
editor: visual
---

## Predicting Students Dropout and Academic Success

Research Question: Does age at enrollment contribute to student academic dropout and failure in higher education?

```{r}
# set up the libraries
library(tidyverse)
library(readxl) 
library(here) # sets path to your folder directory
library(leaps) # for variable selection methods
```

```{r}
# load the data
pj_data <- read_xlsx(here("data", "project_data.xlsx")) # the data file is in a folder named 'data'
head(pj_data)
str(pj_data)
```

```{r}
summary(pj_data)
pj_data <- na.omit(pj_data) # remove any missing values
```

```{r}
# Factorize the categorical variables
# 'Target' is the only categorical variable
col <- c("Target")
pj_data [, col] <- lapply(pj_data[, col], factor)
str(pj_data) # confirm categorization of the column
```

For our research question, "Does age at enrollment contribute to student academic dropout and failure in higher education?", we will take 'Age at Enrollment' as the dependent (Y) variable that we can predict using the other variables as the independent variables.

Let's develop a multivariate linear regression model to predict student academic performance.

Since we have many independent variables, a multivariate linear regression would be ideal. We have 37 variables and therefore, we can 36 (i.e.,37-1) independent variables.

To find a good model that shows this relationship between the dependent and independent variables, we will use the 'forward step wise' variable selection method to select predictor variables that will be good for our model.

```{r}
set.seed(123) # set a pseudorandom number generator
index <- sample(1:nrow(pj_data), 0.3 * nrow(pj_data)) # sample 30% of the data for testing
# split the data
training_data <- pj_data[-index, ]
testing_data <- pj_data[index, ]

model.best <- regsubsets(`Target`~., data = pj_data, nvmax =36)
summary(model.best)

# perfom forward stepwise selection
model.fwd <- regsubsets(`Age at enrollment` ~.,data = pj_data, nvmax = 36, method = "forward") # nvmax = 36 because we have 36 independent variables
summary(model.fwd)
```


```{r}
#Correlation Heatmap
#str(pj_data)

corr_data <- pj_data |>
  select(
    `Application order`,
    `Daytime/evening attendance`,
    `Admission grade`,
    Displaced,
    `Educational special needs`,
    Debtor,
    `Tuition fees up to date`,
    `Age at enrollment`,
    International,
    `Unemployment rate`,
    `Inflation rate`,
    GDP
  )

cor(corr_data)
cor_matrix <- cor(corr_data)
diag(cor_matrix) <- NA

heatmap(cor_matrix, main = "Corr. Matrix Heatmap")

corrplot(cor_matrix, method = "color", type = "upper")

strong_corr <- which(abs(cor_matrix) > 0.7, arr.ind = TRUE)
strong_corr

```

```{r}
# Checking age vs target
ggplot(pj_data, aes(x = `Age at enrollment` , fill= Target))+
  #geom_count()
  #geom_boxplot()
  geom_histogram(position = "identity", alpha = 0.4, bins = 30)
```

```{r}
pj_binned <- pj_data %>%
  mutate(age_bin = cut(`Age at enrollment`,
                       breaks = seq(min(`Age at enrollment`), max(`Age at enrollment`), by = 1),
                       include.lowest = TRUE)) %>%
  group_by(age_bin) %>%
  summarise(
    Dropout = mean(Target == "Dropout"),
    Enrolled = mean(Target == "Enrolled"),
    Graduate = mean(Target == "Graduate"),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(Dropout, Enrolled, Graduate),
               names_to = "Target", values_to = "Probability") %>%
  # Add a numeric column for age using the bin midpoint
  mutate(Age = sapply(strsplit(as.character(age_bin), ","), function(x) {
    bounds <- gsub("\\(|\\]|\\[|\\)", "", x)
    mean(as.numeric(unlist(strsplit(bounds, "-"))))
  }))

# Plot
ggplot(pj_binned, aes(x = Age, y = Probability, color = Target)) +
  geom_smooth() +
  geom_point()+
  labs(
    x = "Age at enrollment",
    y = "Probability (%)",
    title = "Probability curves by age",
    color = "Target"
  ) +
  theme_minimal()

summary(pj_data$`Age at enrollment`)
```